How and Why to run a Hack Week

### 2017-08-17 paper.tex

We note that the terminology for these events is constantly evolving, and that the ''hackathon`` concept may have implicit connotations that are disfavored in some communities. 
One criticism of hackathons is that they propel the ''geek`` stereotype and may present a barrier to creating an inclusive working environment, especially for women \citep{Decker2015}. 
Another problem is the competitive atmosphere of many industry hackathons, where teams actively compete for prizes. 
Because certain groups are more risk-averse than others, and risk-aversion traces demographics like gender, we urge caution when contemplating making an event competitive. 
Other related nomenclature includes the ''unconference`` which is also a time-bounded event gathering together a specific community, but these often have minimal structure and do not necessarily have a focus on software. 
The ''sprint`` or ''scrum`` label typically refers to an event focused on rapid software development on a specific set of code. 
This may be a component of our hack week model but it ignores the important pedagogical components.

@@

\begin{itemize}
\item{\textit{Education and training}: Some hack weeks are more focused on education than others (see Figure 1). 
However, even those that are not explicitly focused on skills-development will include extensive discussions on reproducible and open science practices that are valuable to participants. 
Discussing these topics in a diverse group setting helps to ingrain within participants the basic conventions of the broader open source community. 
Participants can then serve as ''ambassadors`` of these practices and foster their adoption by other researchers in the field. 
Furthermore, the kind of co-working that is inherent to the hack week format facilitates lateral knolwedge transfer, where participants learn from each other about practices and tools of the trade. 
Often these tools are not described in research products such as papers, or even in the software implementations.}
\item{\textit{Tool development}: hack weeks present an opportunity for computational tool developers to get together with users of these tools, and to critically evaluate the use of the tools in addressing particular scientific questions.}

\item{\textit{Community building}: a hack week is an opportunity to galvanize a community within a field that cares about computational issues, and relates to specific open-source software projects. 
For example, because of its focus on free and open-source tools, Neurohack week helps galvanize the Neuroimaging in Python (http://nipy.org) community. 
It also allows researchers who may otherwise be isolated within their departments and subfield on account of their computational specialization to come together and exchange experiences. 
As one participant at Astro Hack Week expressed in the survey, \textit{''The main thing I took from AstroHackWeek is the fact that I'm not alone``}.}

\item{\textit{Interdisciplinary research}: a short time-bounded event is an excellent opportunity to experiment with concepts, questions and methods that traverse boundaries within a discipline (e.g., between researchers using different methods, or studying different systems, or experimental models), as well as across disciplines. 
These experiments are often discouraged in traditional disciplinary research, because of the difficulty to achieve results in both disciplines, and because of the risks inherent in taking a less trodden path\footnote{\url{https://www.ncbi.nlm.nih.gov/labs/articles/12970550} and \url{https://www.researchgate.net/publication/8126355_EDUCATION_Risks_and_Rewards_of_an_Interdisciplinary_Research_Path}}. 
This is despite the fact that interdisciplinary research can also be very impactful \citep{Hall2012-hi}. 
A diversity of research backgrounds, and interactions with scientists from other disciplines, further facilitates these experiments.}

\item{\textit{Recruitment and networking}: hackweeks often bring together research, government and industry communities, providing an opportunity for participants to learn about each other's interests and their abilities in close quarters. 
This kind of extended opportunity for interaction and ''on-the-job`` testing is considered an excellent form of evaluation of prospective job candidates [citation needed]. 
Conversely, participants have an opportunity to gain exposure to potential career opportunities in communities other than their own. 
As an example, one of the participants in Neurohack week 2016 will be joining the eScience Institute as a post-doc in Fall of 2017 (and as an instructor in Neurohack week 2017).}

\item{\textit{It's fun}: the opportunity to focus for a week on a slightly more whimsical and adventurous sort of scientific activities can be both fruitful, and enjoyable. 
Even aside from the prospect of interdisciplinary research, it can provide a laboratory for more experimental approaches and high-risk projects than scientist will take on in their day-to-day research.}

\end{itemize}

It is worth noting that the reasons for participants to attend a hack week are as diverse as the goals that the organizing committee explicitly states. 
This is largely a function of the diversity of the participants (see section below), and should be something that organizers design for. 
Beginner participants may attend primarily to learn a new technique or method, while more experienced participants attend to gain more experience mentoring and teaching. 
Some researchers may come with particular projects and collaborators in mind, while others come with a focus on learning and with no explicit plan for the project-based work. 
While it is very hard to design a workshop that is universally useful to all participants, in practice many researchers attend with the understanding that some components will be more congruent with their own goals than others, but also bring an open mind and a willingness to learn. 
In addition, the diversity of goals can effectively be a strength of the workshop, if the organizers can facilitate matching up participants with complementary goals.

@@

Because hack weeks are at their core non-traditional workshops, it is worth thinking in detail about the target audience for a hack week, along with methods to select an appropriate group of participants. Here, appropriate essentially refers to 'most likely to achieve the stated goal of the hack week', though measuring that achievement remains a difficult issue itself. 
This is especially important because hack weeks, particularly in their more project-heavy iterations, are very participant-driven workshops. 
At a traditional conference or summer school, the program will be largely set in advance by the organizing committee using the careful selection of speakers and lecturers. 
At a hack week much of the program is in flux both before and during the workshop. 
In this context, a successful hack week requires participants who are both interested in participation and feel comfortable participating in discussions and taking risks. 
Therefore, a robust, effective participant selection procedure is of crucial importance to the outcomes of the hack week and may require both careful engineering and an in-depth discussion among the organizers.

Perhaps unsurprisingly, the target audience for a hack week depends directly on the stated goal of the workshop. 
At conferences, researchers typically report recent results to the community and network with other scientists largely from their discipline, leading to a fairly narrow target distribution in academic seniority and represented fields. 
At a summer school, the aim is generally for senior researchers to transfer knowledge about a particular subject or field to novice researchers, primarily graduate students and early postdoctoral fellows. 
Hack weeks differ in that knowledge transfer occurs across many levels of senority, due to the relative newness of the content and the distributed knowledge base. 
Because the potential for knowledge transfer and community building is highest when the group is heterogeneous, participant diverse in both academic background and geographic location should be selected. 
This increases the chances that participants will take their newfound knowledge into their respective communities and spread it further.

Aside from these broad considerations, the tuning of the audience and participant selection will depend directly on several of the axes described in detail above. 
In particular, education-focused workshops might be more homogeneous than project-focused workshops, and similarly single-discipline hack weeks may be more homogeneous than interdisciplinary ones. 
Additional goals may be not directly related to the workshop itself, but on higher-order goals that a community might find desirable and that can be directly addressed by engineering the participant selection.
 For example, women are still a small minority in computational astrophysics. 
 One goal of a hack week could be to increase the participation of gender minorities in computation-focused research. 
 Similarly, if a particular sub-field within a discipline is ahead of others in their computational approaches, it might be worthwhile to engineer the mix of participants such that a number of experts from that sub-field are present. 
 One interesting axis to consider is the mix of academics versus participants from industry. 
 The latter may often be more focused on practically useful applications of methods and data sets, and can therefore provide an important counterpoint to more theory-focused academics. 
 In particular in fields where the career path from academia to industry is less immediately obvious (as this the case e.g. in astronomy, where little directly relevant industry exists), inviting back participants who transitioned to industry successfully can provide important role models and networking opportunities for junior participants. 
 Finally, with much of data science currently performed in the private sector, industry participants can bring directly relevant knowledge and new scientific approaches to the table.

The need to carefully select participants dovetails with another requirement: transparency. 
Up to date, all hack weeks we have organized have been oversubscribed by at least a factor of two, and thus necessarily require turning applicants away. 
The selection procedure should be as transparent to the applicants as possible, and the organizing committee should be held accountable for their performance. 
Transparency is necessary for applicants to understand acceptance/rejection decisions, and accountability is of crucial importance for the detection of inherent biases in the selection, which may harm both the workshop's success as well as the larger community.

@@

In practice, most selection procedures follow two separate steps, though in the case of human committees, these steps are often conflated. 
In the first step, the merit of a candidate must be assessed: does the candidate fulfill the requirements posed by the goals of the workshop? For example, at Astro Hack Week, it may be unrealistic to admit an undergraduate student in biochemistry, unless they had a compelling reason for attending. 
This first step is exceedingly difficult to automate, because it requires judgment calls, often based on long-form answers, that are as of yet difficult for computers to parse and classify. 
In the past, we have performed this step by blinding ourselves to a candidate's other attributes (including name and other personal information), and assess their candidacy based solely upon key questions asked specifically for this purpose.
When doing this procedure for a large enough sample, it is unlikely that the resulting pool of acceptable candidates is smaller than the number of available spaces at the workshop. 
The second step in the selection procedure thus requires tie-breaking between equally acceptable candidates. 
It is here where one may impose outside constraints on the selection based on the goals of the workshop. 
If multiple competing constraints are considered, this task essentially becomes a complex optimization problem, for which provably optimal algorithms exist that will far outperform any human selection procedure.

@@

Within our hack weeks the goal of the selection procedure has always been diversity among several axes. 
This includes academic subdiscipline, to allow for maximally possible knowledge exchange between different domain fields, as well as with researchers from adjacent fields including statistics, computer science and data science. 
Other important categories revolve around the participants' previous knowledge in terms of machine learning, statistics and programming. 
Unlike for a summer school, we have generally selected for a diverse set of participants in order to be able to efficiently pair up novices and experts in projects. 
Finally, we also include demographic information with two main goals: firstly, representation has been proven to be an effective way to foster an inclusive environment where minority participants feel comfortable to participate and take risks. 
Secondly, we would like to boost the participation of minority researchers in computational astrophysics, and our hack week model provides a useful framework for doing so.

@@

To date, all organized hack weeks have been subject-specific. 
That is, they aimed to bring together a community with a shared scientific interest, e.g.\ neuroscience, astronomy or geoscience. 
This has several clear advantages. 
In particular, communities organized by subject generally share the same language, which drastically cuts down on the time participants spend simply agreeing on scientific terms. 
Similarly, the scientific objectives and questions are broadly clear to all participants: there is little need to explain in detail why a particular question may be of interest to the field. 
Finally, there are practical considerations: while even within fields the use of particular data sets and software packages may differ quite widely, there is generally a shared understanding of the type of data usually taken, and the most common ways to analyze this data. 
This leaves more time for active collaboration on cutting-edge science.

However, there are also drawbacks to limiting the theme of a hack week to a single field. 
Homogeneity may be a disadvantage if it leads to \textit{group think} and inhibits new, creative solutions. 
Within a field, the chance of participants being familiar with each other may be larger, leading to the formation of cliques and insufficient mixing. 
These disadvantages may be mitigated by organizing a hack week around a technique (e.g. Gaussian Processes) or modality (e.g. imaging) instead. 
In these case, building a shared vocabulary and understanding the major data analysis problems in each field is a crucial task, and should be allocated sufficient time in form of introductory talks or tutorials to foster cross-disciplinary collaboration. 
On the other hand, cross-disciplinary hack weeks around a specific technique or type of data set allow methods developed in a specific field to diffuse into other subjects and therefore help avoid duplication of method development efforts.

An example of such events are the ImageXD (image processing across domains\footnote{\url{http://http://www.imagexd.org/}} meetings held at the UC Berkeley Institute for Data Science in the Summer 2016, and at the UW eScience Institute in the Spring of 2017. 
These three-day events also included a mix of tutorials, talks, and joint work on projects, but in contrast to the hack weeks described in the present paper, these events included participants from a variety of different research fields, with the common thread being that they all have an interest in the analysis of images, as part their research. 
Thus, audience and speakers included researchers from wide array of domains: astronomy, neuroscience and geosciences, but also material science, chemistry, biology, medicine, engineering, and others. 
In addition, participants included computer vision researchers, who develop algorithms that can be applied to answer scientific questions. 
Similarly, a series of events at UC Berkeley (Text XD) focused on the analysis of text data across different domains of research.

@@

Design considerations depend very strongly on goals. 
In general, longer events allow for a larger taught component, more ambitious projects and especially for cross-disciplinary events are more likely to provide enough time for different groups to effectively communicate across barriers of professional terminology. 
On the other hand, because of their large participatory component, hack weeks tend to be exhausting, and long events may lead to fatigue among attendees, where active participation and measurable outcomes may drop sharply in the latter days of the workshops. 
This is particularly important if---as is often the case---participants continue participating in workshop-relevant activities even after the official component ends for the day.

Ways to mitigate these effects include capping taught components during the day, providing a reasonable clear schedule, and limiting parallel components to avoid decision fatigue. 
At Astro Hack Week, we have limited the scheduled taught components to no more than three out of nine hours for each of the five workshop days, and unscheduled (shorter) tutorials to a maximum of two per day. 
Even so, in our surveys, participants occasionally report concerns about choosing to attend tutorials versus working on projects.

Aside from time, space plays a particularly important role in facilitating a successful hack week. 
Universities in general provide a convenient venue with existing structures to facilitate hack weeks (e.g. access to scientific publications, institutional support including staff and funding). 
On the other hand, most spaces in universities are designed for lectures, which are diametrically opposed to unconference-style events. 
Thus, finding an appropriate physical space within a university may be a challenge. 
However, with the rise of active learning as a preferred teaching methods, traditional lecture spaces are transforming into more flexible spaces that are generally appropriate for hack weeks. 
As a general rule, the smaller a hack week's emphasis on a taught component, the more flexible the space has to be, with ample opportunity for re-configuration. 
At Astro Hack Week, we have found it beneficial to spend one day mid-week at a different location (e.g. a company headquarter) to engage participants and break the routine. 
It is worth noting that collaborative spaces for non-traditional workshops are becoming more prevalent in academia (some attached to universities, some not) and may provide support and infrastructure that a traditional university location may not.

Another important design consideration is group size. 
With all previous hack weeks severely oversubscribed, it seems natural to simply admit more participants. 
However, this can counter the ideals of the workshop, in particular when building a community is one of its stated goals. 
If the group is too large, participants are unlikely to even meet each other, and workshop cohesion may be lost as the workshop fractures into smaller groups, often among participants who already know each other. 
This may inhibit knowledge transfer by clustering participants into small in-groups. 
Additionally, it is likely that the number participant-led components in the schedule may increase with workshop size. 
While generally desirable, a programme that is too crowded may lead to fatigue On the other hand, if the size of the workshop is too small, it is unlikely to achieve the desired diversity among participants to foster new collaborations across sub-fields and disciplines. 
In the past, we have found groups with sizes between 50 and 70 participants to be large enough to encourage a breadth of projects while allowing the workshop to function as a cohesive group.

As mentioned above, the balance between pedagogy and working depends both on the goals of the workshop and the topics around which the workshop is organized. 
If participants have little shared knowledge, more teaching may be necessary in order to allow participants to effectively communicate with each other. 
In communities where a shared understanding exists, tutorials can focus on more advanced or innovative topics, and less time may be allocated for them, leaving more time for active participation. 
In astronomy, a relatively small field, even students generally share a common knowledge base and have rudimentary knowledge about the types of data used and the challenges associated with each. 
Thus, we have focused the taught component of Astro Hack Week less on domain-specific knowledge, and instead offered tutorials in topics from domain-adjacent fields like statistics and computer science that attendees are unlikely to have encountered in their regular education.

Hack week outcomes, in turn, depend strongly on participants and are often a function of their interests and seniority. 
A number of attendees arrive with the stated goal of writing a specific scientific article, though these tend to be more advanced participants with significant pre-knowledge of both hackathons and their topic of interest. 
Conversely, many attendees arrive with the stated goal of learning about a specific topic (such as machine learning) and bring a specific data set they believe the new knowledge may be applicable to. 
This leads to a wide variety of project types from sandbox-style explorations to focused work efforts. 
It is worth noting that while a scientific paper need not be the stated goal of a hack (and is unlikely to be completed in the short time allocated in any case), results may still be published as short reports or unconference proceedings. 
For example, Neurohack week provides a venue for participants to publish a short (two-page) ''project report`` summarizing the hack that participants did during the week of NHW. 
Similarly, Python in Astronomy gathers all documents produced during the workshop (unconference transcripts, talk summaries, descriptions of sprint and hack projects) into citeable unconference proceedings.

@@

Measuring the success of a hack week objectively is complicated by the diversity of goals of both organizing committee and attendees. 
A participant who attends primarily in order to hack on a specific project may find the tutorials less overall useful than a graduate student exposed to the tutorial topics for the first time. 
Additionally, the participant-driven, open-ended format facilitates knowledge transfer and collaborations in sometimes surprising ways that escape traditional measures of success. 
Additionally, it is difficult to define in the first place what success means to organizers and success. 
For more project-driven events, one key metric is the number of publications that result from hack week projects. 
However, this implies a very narrow definition of success in line with standard academic performance indicators, and may be hard to collect in practice, if participants do not explicitly include a reference to the event in their publication. 
Another, somewhat less traditional and complementary indicator is the activity of participants in terms of code written and committed to a public code repository (e.g. GitHub). 
This could be defined in lines of code committed, the number of new open-source projects spawned, and the number of new developers contributing to open-source projects. 
This assumes that participants work largely in the open during a hack week, and that most projects have a strong programming component. 
Again, this may not be true for some hacks (e.g. tutorials, other written text, visualizations), and focuses entirely on material outcomes. 
In particular, it ignores learning, community building as well as networking outcomes, which are often equally important objectives. 
One possible approach here is to ask participants to self-report their success via a post-workshop survey.
If participants are selected as described above, it is in principle possible to measure gains in knowledge, networks and productivity within the pool of acceptable candidates for both those that attended the hack week and those that did not. 
This would then provide a somewhat objective measure of the impact the hack week has had. 
In practice, this has not been done for any of the hack weeks conducted thus far.

For practical purposes, we recommend an approach that combines all of these metrics: publications directly resulting out of hack week collaborations, new code generated and new contributors to open-source projects, self-reported learning outcomes in the topics covered in the tutorials, and survey results probing the usefulness of the workshop in terms of career and networking goals. 
Open-ended questions allow participants to provide feedback about outcomes, problems and goals that organizers had not anticipated.

>>>>

We note that the terminology for these events is constantly evolving, and that the ''hackathon`` concept may have implicit connotations that are disfavored in some communities. 
One criticism of hackathons is that they propel the ''geek`` stereotype and may present a barrier to creating an inclusive working environment, especially for women \citep{Decker2015}. 
Another problem is the competitive atmosphere of many industry hackathons, where teams actively compete for prizes. 
Because certain groups are more risk-averse than others, and risk-aversion traces demographics like gender, we urge caution when contemplating making an event competitive. 
Other related nomenclature includes the ''unconference`` which is also a time-bounded event gathering together a specific community, but these often have minimal structure and do not necessarily have a focus on software. 
The ''sprint`` or ''scrum`` label typically refers to an event focused on rapid software development on a specific set of code. 
%This may be a component of our hack week model but it ignores the important pedagogical components.

@@

\begin{itemize}
\item{\textit{Education and training}: %Some hack weeks are more focused on education than others (see Figure 1). 
Even those hackathons not explicitly focused on skills-development will include extensive discussions on reproducible and open science practices that are valuable to participants. 
%Discussing these topics in a diverse group setting helps to ingrain within participants the basic conventions of the broader open source community. 
Participants can then serve as ''ambassadors`` of these practices and foster their adoption by other researchers in the field. 
Furthermore, the kind of co-working that is inherent to the hack week format facilitates lateral knowledge transfer, where participants learn from each other about practices and tools of the trade.}
%Often these tools are not described in research products such as papers, or even in the software implementations.}
\item{\textit{Tool development}: hack weeks present an opportunity for computational tool developers to get together with users of these tools, and to critically evaluate the use of the tools in addressing particular scientific questions.}
\item{\textit{Community building}: a hack week is an opportunity to galvanize a community within a field.% that cares about computational issues, and relates to specific open-source software projects. 
For example, because of its focus on free and open-source tools, Neurohack week helps galvanize the Neuroimaging in Python (http://nipy.org) community. 
It also allows researchers who may otherwise be isolated within their departments and subfield on account of their computational specialization to come together and exchange experiences.}
%As one participant at Astro Hack Week expressed in the survey, \textit{''The main thing I took from Astrohack week is the fact that I'm not alone``}.}
\item{\textit{Interdisciplinary research}: a short time-bounded event is an excellent opportunity to experiment with concepts, questions and methods that traverse boundaries within a discipline (e.g.\ between researchers using different methods, or studying different systems, or experimental models), as well as across disciplines. 
Despite the impact of interdisciplinary research \citep{Hall2012-hi}, these experiments are often discouraged in traditional disciplinary research, because of the difficulty to achieve results in both disciplines, and because of the risks inherent in taking a less trodden path\footnote{\url{https://www.ncbi.nlm.nih.gov/labs/articles/12970550} and \url{https://www.researchgate.net/publication/8126355_EDUCATION_Risks_and_Rewards_of_an_Interdisciplinary_Research_Path}}.}
%This is despite the fact that interdisciplinary research can also be very impactful \citep{Hall2012-hi}. 
%A diversity of research backgrounds, and interactions with scientists from other disciplines, further facilitates these experiments.}

\item{\textit{Recruitment and networking}: hack weeks often bring together research, government and industry communities, providing an opportunity for participants to learn about each other's interests and their abilities in close quarters. 
This kind of extended opportunity for interaction and ''on-the-job`` testing is considered an excellent form of evaluation of prospective job candidates [CITATION NEEDED]. 
Conversely, participants have an opportunity to gain exposure to potential career opportunities in communities other than their own.}
%As an example, one of the participants in Neurohack week 2016 will be joining the eScience Institute as a post-doc in Fall of 2017 (and as an instructor in Neurohack week 2017).}
%\item{\textit{It's fun}: the opportunity to focus for a week on a slightly more whimsical and adventurous sort of scientific activities can be both fruitful, and enjoyable. 
%Even aside from the prospect of interdisciplinary research, it can provide a laboratory for more experimental approaches and high-risk projects than scientist will take on in their day-to-day research.}
\end{itemize}

It is worth noting that the reasons for participants to attend a hack week are as diverse as the goals that the organizing committee explicitly states. 
This is largely a function of the diversity of the participants (see section below), and should be something that organizers design for. 
Beginner participants may attend primarily to learn a new technique or method, while more experienced participants attend to gain more experience mentoring and teaching. 
Some researchers may come with particular projects and collaborators in mind, while others come with a focus on learning and with no explicit plan for the project-based work. 
%While it is very hard to design a workshop that is universally useful to all participants, in practice many researchers attend with the understanding that some components will be more congruent with their own goals than others, but also bring an open mind and a willingness to learn. 
%In addition, the diversity of goals can effectively be a strength of the workshop, if the organizers can facilitate matching up participants with complementary goals.

@@

Because hack weeks are at their core non-traditional workshops, it is worth thinking in detail about the target audience for a hack week, along with methods to select an appropriate group of participants. Here, appropriate essentially refers to 'most likely to achieve the stated goal of the hack week', though measuring that achievement remains a difficult issue itself. 
This is especially important because hack weeks are very participant-driven workshops. 
%At a traditional conference or summer school, the program will be largely set in advance by the organizing committee using the careful selection of speakers and lecturers. 
%At a hack week much of the program is in flux both before and during the workshop. 
In this context, a successful hack week requires participants who are both interested in participation and feel comfortable participating in discussions and taking risks. 
Therefore, a robust, effective participant selection procedure is of crucial importance to the outcomes of the hack week and may require both careful engineering and an in-depth discussion among the organizers.

Perhaps unsurprisingly, the target audience for a hack week depends directly on the stated goal of the workshop. 
At conferences, researchers typically report recent results to the community and network with other scientists largely from their discipline, leading to a fairly narrow target distribution in academic seniority and represented fields. 
At a summer school, the aim is generally for senior researchers to transfer knowledge about a particular subject or field to novice researchers, primarily graduate students and early postdoctoral fellows. 
Hack weeks differ in that knowledge transfer occurs across many levels of seniority, due to the relative newness of the content and the distributed knowledge base. 
%Because the potential for knowledge transfer and community building is highest when the group is heterogeneous, participant diverse in both academic background and geographic location should be selected. 
%This increases the chances that participants will take their newfound knowledge into their respective communities and spread it further.

Aside from these broad considerations, the tuning of the audience and participant selection will depend directly on several of the axes described in detail above. 
In particular, education-focused workshops might be more homogeneous than project-focused workshops, and similarly single-discipline hack weeks may be more homogeneous than interdisciplinary ones. 
Additional goals may be not directly related to the workshop itself, but on higher-order goals (e.g.\ increasing minority participation, knowledge transfer between academia and industry) that a community might find desirable and that can be directly addressed by engineering the participant selection.
% For example, women are still a small minority in computational astrophysics. 
%One goal of a hack week could be to increase the participation of gender minorities in computation-focused research. 
%Similarly, if a particular sub-field within a discipline is ahead of others in their computational approaches, it might be worthwhile to engineer the mix of participants such that a number of experts from that sub-field are present. 
%One interesting axis to consider is the mix of academics versus participants from industry. 
%The latter may often be more focused on practically useful applications of methods and data sets, and can therefore provide an important counterpoint to more theory-focused academics. 
%In particular in fields where the career path from academia to industry is less immediately obvious (as this the case e.g. in astronomy, where little directly relevant industry exists), inviting back participants who transitioned to industry successfully can provide important role models and networking opportunities for junior participants. 
%Finally, with much of data science currently performed in the private sector, industry participants can bring directly relevant knowledge and new scientific approaches to the table.

The need to carefully select participants dovetails with another requirement: transparency. 
%Up to date, all hack weeks we have organized have been oversubscribed by at least a factor of two, and thus necessarily require turning applicants away. 
The selection procedure should be as transparent to the applicants as possible, and the organizing committee should be held accountable for their performance. 
Transparency is necessary for applicants to understand acceptance/rejection decisions, and accountability is of crucial importance for the detection of inherent biases in the selection, which may harm both the workshop's success as well as the larger community.

@@

In practice, most selection procedures follow two separate steps, though in the case of human committees, these steps are often conflated. 
In the first step, the merit of a candidate must be assessed: does the candidate fulfill the requirements posed by the goals of the workshop? %For example, at Astro Hack Week, it may be unrealistic to admit an undergraduate student in biochemistry, unless they had a compelling reason for attending. 
This first step is difficult to automate, because it requires judgment calls, often based on long-form answers. 
In the past, we have performed this step by blinding ourselves to a candidate's other attributes (including name and other personal information), and assess their candidacy based solely upon key questions asked specifically for this purpose.
When doing this procedure for a large enough sample, it is unlikely that the resulting pool of acceptable candidates is smaller than the number of available spaces at the workshop. 
The second step in the selection procedure thus requires tie-breaking between equally acceptable candidates. 
It is here where one may impose outside constraints on the selection based on the goals of the workshop. 
If multiple competing constraints are considered, this task essentially becomes a complex optimization problem, for which algorithms exist that will outperform any human selection procedure.

@@

%Within our hack weeks the goal of the selection procedure has always been diversity among several axes. 
%This includes academic subdiscipline, to allow for maximally possible knowledge exchange between different domain fields, as well as with researchers from adjacent fields including statistics, computer science and data science. 
%Other important categories revolve around the participants' previous knowledge in terms of machine learning, statistics and programming. 
%Unlike for a summer school, we have generally selected for a diverse set of participants in order to be able to efficiently pair up novices and experts in projects. 
%Finally, we also include demographic information with two main goals: firstly, representation has been proven to be an effective way to foster an inclusive environment where minority participants feel comfortable to participate and take risks. 
%Secondly, we would like to boost the participation of minority researchers in computational astrophysics, and our hack week model provides a useful framework for doing so.

@@

To date, all organized hack weeks have been subject-specific. 
That is, they aimed to bring together a community with a shared scientific interest, e.g.\ neuroscience, astronomy or geoscience. 
This has several clear advantages. 
In particular, communities organized by subject generally share the same language.%, which drastically cuts down on the time participants spend simply agreeing on scientific terms. 
Similarly, the scientific objectives and questions are broadly clear to all participants.%: there is little need to explain in detail why a particular question may be of interest to the field. 
Finally, there are practical considerations: while even within fields the use of particular data sets and software packages may differ quite widely, there is generally a shared understanding of the type of data usually taken, and the most common ways to analyze this data. 
This leaves more time for active collaboration on cutting-edge science.

However, there are also drawbacks to limiting the theme of a hack week to a single field. 
Homogeneity may be a disadvantage if it leads to \textit{group think} and inhibits new, creative solutions. 
Within a field, the chance of participants being familiar with each other may be larger, leading to the formation of cliques and insufficient mixing. 
These disadvantages may be mitigated by organizing a hack week around a technique (e.g. Gaussian Processes) or modality (e.g. imaging) instead. 
In these case, building a shared vocabulary and understanding the major data analysis problems in each field is a crucial task.%, and should be allocated sufficient time in form of introductory talks or tutorials to foster cross-disciplinary collaboration. 
On the other hand, cross-disciplinary hack weeks around a specific technique or type of data set allow methods developed in a specific field to diffuse into other subjects and therefore help avoid duplication of method development efforts.

An example of such events are the ImageXD (image processing across domains\footnote{\url{http://http://www.imagexd.org/}} meetings held at the UC Berkeley Institute for Data Science in the Summer 2016, and at the UW eScience Institute in the Spring of 2017. 
These three-day events also included a mix of tutorials, talks, and joint work on projects, but in contrast to the hack weeks described in the present paper, these events included participants from a variety of different research fields, with the common thread being that they all have an interest in the analysis of images, as part their research. 
%Thus, audience and speakers included researchers from wide array of domains: astronomy, neuroscience and geosciences, but also material science, chemistry, biology, medicine, engineering, and others. 
%In addition, participants included computer vision researchers, who develop algorithms that can be applied to answer scientific questions. 
%Similarly, a series of events at UC Berkeley (Text XD) focused on the analysis of text data across different domains of research.

@@

Design considerations depend very strongly on goals. 
In general, longer events allow for a larger taught component, more ambitious projects and especially for cross-disciplinary events are more likely to provide enough time for different groups to effectively communicate across barriers of professional terminology. 
On the other hand, because of their large participatory component, hack weeks tend to be exhausting, and long events may lead to fatigue among attendees, where active participation and measurable outcomes may drop sharply in the latter days of the workshops. 
%This is particularly important if---as is often the case---participants continue participating in workshop-relevant activities even after the official component ends for the day.
Ways to mitigate these effects include capping taught components during the day, providing a reasonable clear schedule, and limiting parallel components to avoid decision fatigue. 
%At Astro Hack Week, we have limited the scheduled taught components to no more than three out of nine hours for each of the five workshop days, and unscheduled (shorter) tutorials to a maximum of two per day. 
%Even so, in our surveys, participants occasionally report concerns about choosing to attend tutorials versus working on projects.

Aside from time, space plays a particularly important role in facilitating a successful hack week. 
Universities in general provide a convenient venue with existing structures to facilitate hack weeks (e.g. access to scientific publications, institutional support including staff and funding). 
On the other hand, most spaces in universities are designed for lectures, which are diametrically opposed to unconference-style events. 
Thus, finding an appropriate physical space within a university may be a challenge. 
However, with the rise of active learning as a preferred teaching methods, traditional lecture spaces are transforming into more flexible spaces that are generally appropriate for hack weeks. 
As a general rule, the smaller a hack week's emphasis on a taught component, the more flexible the space has to be, with ample opportunity for re-configuration. 
%At Astro Hack Week, we have found it beneficial to spend one day mid-week at a different location (e.g. a company headquarter) to engage participants and break the routine. 
%It is worth noting that collaborative spaces for non-traditional workshops are becoming more prevalent in academia (some attached to universities, some not) and may provide support and infrastructure that a traditional university location may not.

Another important design consideration is group size. 
%With all previous hack weeks severely oversubscribed, it seems natural to simply admit more participants. 
%However, this can counter the ideals of the workshop, in particular when building a community is one of its stated goals. 
If the group is too large, participants are unlikely to even meet each other, and workshop cohesion may be lost as the workshop fractures into smaller groups, often among participants who already know each other. 
This may inhibit knowledge transfer by clustering participants into small in-groups. 
%Additionally, it is likely that the number participant-led components in the schedule may increase with workshop size. 
%While generally desirable, a programme that is too crowded may lead to fatigue 
On the other hand, if the size of the workshop is too small, it is unlikely to achieve the desired diversity among participants to foster new collaborations across sub-fields and disciplines. 
In the past, we have found groups with sizes between 50 and 70 participants to be large enough to encourage a breadth of projects while allowing the workshop to function as a cohesive group.

As mentioned above, the balance between pedagogy and working depends both on the goals of the workshop and the topics around which the workshop is organized. 
If participants have little shared knowledge, more teaching may be necessary in order to allow participants to effectively communicate with each other. 
In communities where a shared understanding exists, tutorials can focus on more advanced or innovative topics, and less time may be allocated for them, leaving more time for active participation. 
%In astronomy, a relatively small field, even students generally share a common knowledge base and have rudimentary knowledge about the types of data used and the challenges associated with each. 
%Thus, we have focused the taught component of Astro Hack Week less on domain-specific knowledge, and instead offered tutorials in topics from domain-adjacent fields like statistics and computer science that attendees are unlikely to have encountered in their regular education.

Hack week outcomes, in turn, depend strongly on participants and are often a function of their interests and seniority. 
some attendees arrive with the stated goal of writing a specific scientific article, usually more advanced participants with significant pre-knowledge of both hackathons and their topic of interest. 
Many attendees arrive with the plan to learn a specific topic (such as machine learning) or bring a specific data set they believe the new knowledge may be applicable to. 
This leads to a wide variety of project types from sandbox-style explorations to focused work efforts. 
%It is worth noting that while a scientific paper need not be the stated goal of a hack (and is unlikely to be completed in the short time allocated in any case), results may still be published as short reports or unconference proceedings. 
%For example, Neurohack week provides a venue for participants to publish a short (two-page) ''project report`` summarizing the hack that participants did during the week of NHW. 
%Similarly, Python in Astronomy gathers all documents produced during the workshop (unconference transcripts, talk summaries, descriptions of sprint and hack projects) into citeable unconference proceedings.

@@

Measuring the success of a hack week objectively is complicated by the diversity of goals of both organizing committee and attendees. 
A participant who attends primarily in order to hack on a specific project may find the tutorials less overall useful than a graduate student exposed to the tutorial topics for the first time. 
Additionally, the participant-driven, open-ended format facilitates knowledge transfer and collaborations in sometimes surprising ways that escape traditional measures of success. 
For more project-driven events, one key metric is the number of publications that result from hack week projects. 
However, this implies a very narrow definition of success in line with standard academic performance indicators, and may be hard to collect in practice, if participants do not explicitly include a reference to the event in their publication. 
Another, somewhat less traditional and complementary indicator is the activity of participants in terms of code written and committed to a public code repository (e.g. GitHub). 
This could be defined in lines of code committed, the number of new open-source projects spawned, and the number of new developers contributing to open-source projects. 
This assumes that participants work largely in the open during a hack week, and that most projects have a strong programming component. 
Again, this may not be true for some hacks (e.g. tutorials, other written text, visualizations), and focuses entirely on material outcomes. 
These measures ignore learning, community building as well as networking outcomes, which are often equally important objectives. 
One possible approach here is to ask participants to self-report their success via a post-workshop survey.
It is in principle possible to measure gains in knowledge, networks and productivity within the pool of acceptable candidates for both those that attended the hack week and those that did not. 
%This would then provide a somewhat objective measure of the impact the hack week has had. 
%In practice, this has not been done for any of the hack weeks conducted thus far.
For practical purposes, we recommend an approach that combines all of these metrics: publications, new code and projects generated, self-reported learning outcomes, and survey results. 
%Open-ended questions allow participants to provide feedback about outcomes, problems and goals that organizers had not anticipated.

If hack week organizers plan to conduct research that involves hack week participants (for example, using assessments and evaluations in a paper such as this one) it is important to obtain approval from the Institutional Review Board (IRB), or equivalent body that approves research with human participants at the institution hosting the hack week. 
Though this research would usually fall under the category of ''minimal risk``, it is still important to establish procedures to manage these data, and to obtain informed consent. 
In particular, in some cases hack week organizers may be interested in studying not only the participants in the hack week, but also the applicants who did not end up participating. 
%It is important to establish procedures to conduct such studies and to obtain approval from an IRB.
